Memory for television episodes preserves event content while introducing new across-event similarities

### 2019-12-11 paper/main.tex
updated text stats, removed placeholder fig

We can also apply two list-learning-native analyses that describe how participants group items in their recall sequences: temporal clustering and semantic clustering \citep[][see \textit{Methods} for details]{PolyEtal09}.  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed video events were clustered heavily in participants' recall event sequences (mean: 0.767, SEM: 0.029), and that participants with higher temporal clustering scores also tended to perform better according to both \cite{ChenEtal17}'s hand-annotated memory scores (Pearson's $r(15) = 0.62, p = 0.008$) and our model's estimate (Pearson's $r(15) = 0.54, p = 0.024$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar video events together (mean: 0.787, SEM: 0.018), and participants who exhibited stronger semantic clustering remembered more video events overall (Pearson's $r(15) = 0.55, p = 0.02$).

% Overall, we found that participants heavily clustered video events in their recalls by both temporal proximity (mean: 0.860, SEM: 0.016) and semantic content (mean: 0.888, SEM: 0.014).

% Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.

% ------ NOTE: ------
% Temporal & semantic clustering scores' correlations with number of events & model-estimated memory are no longer significant.  Worth mentioning at all then?
% temporal x hand-annotated: r=.5, p=0.042
% temporal x model-estimated: r=.36, p=.16
% temporal x n events: r=.29, p=.26
% semantic x hand-annotated: r=.32, p=.21
% semantic x model-estimated: r=.33, p=.20
% semantic x n events: r=.48, p=.053
% - Paxton

@@

Statistical models of memory studies often treat memory recalls as binary (e.g. the item was recalled or not) and independent events.  However, our framework produces a content-based model of individual stimulus and recall events, allowing for direct quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed two novel metrics for quantifying naturalistic memory representations:~\textit{precision} and \textit{distinctiveness}.  We define precision as the average correlation between the topic proportions of each recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that, across participants, a higher precision score is correlated to both hand-annotated memory performance (Pearson's $r(15) = 0.55, p = 0.021$) and the number of recall events estimated by our model (Pearson's $r(15) = 0.66, p = 0.004$). A second novel metric we introduce here is distinctiveness, or how unique the recall description was to each video event. We define distinctiveness as 1 minus the average of all non-matching recall events from the video-recall correlation matrix. We hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  Similarly to precision, we find that the more distinct participants recalls are (on average), the more they remembered (hand-annotated memory: Pearson's $r(15) = 0.62, p = 0.007$; number of events: Pearson's $r(15) = 0.78, p < 0.001$).  In summary, using two novel metrics afforded by our approach, we find that participants whose recalls are both more precise and distinct remember more content.
% ------ NOTE: ------
% Correlations between precision/distinctiveness and number of events need to be updated once analysis is re-run
% - Paxton

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} A video-recall correlation matrix for a representative participant (17).  The yellow boxes highlight the maximum correlation in each column.  Precision was computed as the average of the maximum correlation in each column.  On the other hand, distinctiveness was defined as the average of everything except for the maximum correlation in each column. \textbf{B.} The (Pearson's) correlation between precision and hand-annotated memory performance. \textbf{C.} The correlation between precision and the number of events recovered by the model ($k$). \textbf{D.} The correlation between distinctiveness and hand-annotated memory performance. \textbf{E.} The correlation between distinctiveness and the number of events recovered by the model ($k$).}
  \label{fig:precision-distinctiveness}
\end{figure}
% ------ NOTE: ------
% Figure and caption need to be updated once analysis is re-run
% - Paxton

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/precision_distinctiveness_example}
\caption{\small \textbf{Precision and Distinctiveness Example.}}
\label{fig:precision-distinctiveness-example}
\end{figure}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/trajectory}
\caption{\small \textbf{Trajectories through topic space capture the dynamic content of the video and recalls.}  All panels: the topic proportion matrices have been projected onto a shared two-dimensional space using UMAP.  \textbf{A.} The two-dimensional topic trajectory taken by the episode of \textit{Sherlock}.  Each dot indicates an event identified using the HMM (see \textit{Methods}); the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B.} The average two-dimensional trajectory captured by participants' recall sequences, with the same format and coloring as the trajectory in Panel A.  To compute the event positions, we matched each recalled event with an event from the original video (see \textit{Results}), and then we averaged the positions of all events with the same label.  The arrows reflect the average transition direction through topic space taken by any participants whose trajectories crossed that part of topic space; blue denotes reliable agreement across participants via a Rayleigh test ($p < 0.05$, corrected).  \textbf{C.} The recall topic trajectories (gray) taken by each individual participant (P1--P17).  The video's trajectory is shown in black for reference.  (Same format and coloring as Panel A.)}
\label{fig:trajectory}
\end{figure}

>>>>

We can also apply two list-learning-native analyses that describe how participants group items in their recall sequences: temporal clustering and semantic clustering \citep[][see \textit{Methods} for details]{PolyEtal09}.  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed video events were clustered heavily in participants' recall event sequences (mean: 0.767, SEM: 0.029), and that participants with higher temporal clustering scores tended to perform better according to both \cite{ChenEtal17}'s hand-annotated memory scores (Pearson's $r(15) = 0.62,~p = 0.008$) and our model's estimate (Pearson's $r(15) = 0.54,~p = 0.024$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar video events together (mean: 0.787, SEM: 0.018), and that semantic clustering score was also related to both hand-annotated  (Pearson's $r(15) = 0.65,~p = 0.004$) and model-derived (Pearson's $r(15) = 0.63,~p = 0.007$) memory performance.

% NOTE: 
% Semantic and temporal clustering are both positively correlated with n_events -- r=0.49; p=0.046 and r=0.61; p=0.009, respectively.  
% - Paxton

@@

Statistical models of memory studies often treat memory recalls as binary (e.g. the item was recalled or not) and independent events.  However, our framework produces a content-based model of individual stimulus and recall events, allowing for direct quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed two novel metrics for quantifying naturalistic memory representations:~\textit{precision} and \textit{distinctiveness}.  We define precision as the average correlation between the topic proportions of each recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that, across participants, a higher precision score is correlated to both hand-annotated memory performance (Pearson's $r(15) = 0.56, p = 0.021$) and the number of recall events estimated by our model (Pearson's $r(15) = 0.85, p < 0.001$). A second novel metric we introduce here is distinctiveness, or how unique the recall description was to each video event. We define distinctiveness as 1 minus the average of all non-matching recall events from the video-recall correlation matrix. We hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  We find that this distinctiveness score is related to our model's estimated number of recalled events (Pearson's $r(15) = 0.49, p = 0.046$) but not to the analogous hand-annotated metric (Pearson's $r(15) = 0.31, p = 0.23$).  In summary, using two novel metrics afforded by our approach, we find that participants whose recalls are both more precise and distinct remember more content.
% ------ NOTE: ------
% rephrase/reframe distinctiveness?
% - Paxton

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} A video-recall correlation matrix for a representative participant (17).  The yellow boxes highlight the maximum correlation in each column.  Precision was computed as the average of the maximum correlation in each column.  On the other hand, distinctiveness was defined as the average of everything except for the maximum correlation in each column. \textbf{B.} The (Pearson's) correlation between precision and hand-annotated memory performance. \textbf{C.} The correlation between precision and the number of events recovered by the model ($k$). \textbf{D.} The correlation between distinctiveness and hand-annotated memory performance. \textbf{E.} The correlation between distinctiveness and the number of events recovered by the model ($k$).}
  \label{fig:precision-distinctiveness}
\end{figure}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/trajectory}
\caption{\small \textbf{Trajectories through topic space capture the dynamic content of the video and recalls.}  All panels: the topic proportion matrices have been projected onto a shared two-dimensional space using UMAP.  \textbf{A.} The two-dimensional topic trajectory taken by the episode of \textit{Sherlock}.  Each dot indicates an event identified using the HMM (see \textit{Methods}); the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B.} The average two-dimensional trajectory captured by participants' recall sequences, with the same format and coloring as the trajectory in Panel A.  To compute the event positions, we matched each recalled event with an event from the original video (see \textit{Results}), and then we averaged the positions of all events with the same label.  The arrows reflect the average transition direction through topic space taken by any participants whose trajectories crossed that part of topic space; blue denotes reliable agreement across participants via a Rayleigh test ($p < 0.05$, corrected).  \textbf{C.} The recall topic trajectories (gray) taken by each individual participant (P1--P17).  The video's trajectory is shown in black for reference.  (Same format and coloring as Panel A.)}
\label{fig:trajectory}
\end{figure}
% ------ NOTE: ------
% will update trajectory fig when new version is finished
% - Paxton
